## Loading and preprocessing the data  
```{r,echo=TRUE}
# setwd("~/Downloads/RepData_PeerAssessment1")
# disabled because there are better alternative (download.file) which will easen for non-technical people
# and can be run from any working directory
# download file from repository link
download.file("https://d396qusza40orc.cloudfront.net/repdata/data/activity.zip",destfile="activity.zip")
# unzip and read the downloaded file into R
raw.data<-read.csv(unz("activity.zip","activity.csv"),sep=",",na.strings="NA",stringsAsFactors=FALSE)
# strptime date from date column
date<-strptime(raw.data$date,format="%Y-%m-%d")
# make the interval value more uniform in number format
interval<-formatC(raw.data$interval,width=4,flag="0")
# construct date.time which contain date and time interval value in POSIXlt and POSIXct
date.time<-strptime(paste(raw.data$date,interval),format="%Y-%m-%d %H%M")
```

## What is mean total number of steps taken per day?  
1. Calculate the total number of steps taken per day    
2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day     
3. Calculate and report the mean and median of the total number of steps taken per day  
```{r ,echo=TRUE}
# calculate for total number of steps per day
tot.steps.date<-as.numeric(tapply(raw.data$steps,raw.data$date,sum,na.rm=TRUE))
# plot histogram of total number of steps oer day
hist(tot.steps.date,main="Histogram of Total Number of Steps per Day",xlab="Total Steps in a Day",ylab="Frequency")
# calculate mean of total numbers of steps taken per days
mean.steps.date<-mean(tot.steps.date,na.rm=TRUE)
mean.steps.date
# calculate median of total numbers of steps taken per days
median.steps.date<-median(tot.steps.date,na.rm=TRUE)
median.steps.date
```

## What is the average daily activity pattern?  
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
```{r,echo=TRUE}
# averaging number of the 5 minute interval (x axis), and steps taken, averaged across all day(y axis)
interval.steps.date<-as.numeric(lapply(split(raw.data$steps,raw.data$interval),mean,na.rm=TRUE))
# construct a plot of interval and average steps taken across all days
plot(unique((interval)),interval.steps.date,type="l",main="Average Daily Number of Steps Taken in 5 Minutes Interval",xlab="Interval",ylab="Average of Number of Steps Taken")
# find 5 minute interval across all the days in the data sets  that contain the max number of steps
interval[which.max(interval.steps.date)]
```

## Imputing missing values  
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.  
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)  
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.  
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?  
```{r,echo=TRUE}
# calculate and report number of missing values in the data sets
colSums(is.na(raw.data))
# calculate filling values
# means of steps taken for the corresponding day
fill.mean.date<-(tapply(raw.data$steps,raw.data$date,mean,na.rm=TRUE))
# medians of steps taken for the corresponding day
fill.median.date<-(tapply(raw.data$steps,raw.data$date,median,na.rm=TRUE))
# means of steps taken for the corresponding interval
fill.mean.interval<-(tapply(raw.data$steps,raw.data$interval,mean,na.rm=TRUE))
# medians of steps taken for the corresponding interval
fill.median.interval<-(tapply(raw.data$steps,raw.data$interval,median,na.rm=TRUE))
# check for each filling strategy
fill.mean.date
fill.median.date
fill.mean.interval
fill.median.interval
# READ THE ASSUMPTION MADE: fill.mean.interval is used as filling value
# we chooses fill.mean.interval as filling value
fill.data<-raw.data
fill.data[is.na(fill.data)]<-round(fill.mean.interval)
# calculate for total number of steps per day of filled dataset
fill.tot.steps.date<-as.numeric(tapply(fill.data$steps,fill.data$date,sum,na.rm=TRUE))
#compare the value with the version where na is omitted
cbind(tot.steps.date,fill.tot.steps.date)
# construct histogram of total number of steps taken each day of filled dataset
hist(fill.tot.steps.date,main="Histogram of Total Number of Steps per Day - NAs Replaced'",xlab="Total Steps in a Day",ylab="Frequency")
# mean of total number of steps taken each of filled dataset
fill.mean.steps.date<-mean(fill.tot.steps.date,na.rm=TRUE)
#compare the value with the version where na is omitted
c(mean.steps.date,fill.mean.steps.date)
# median of total number of steps taken each of filled dataset
fill.median.steps.date<-median(fill.tot.steps.date,na.rm=TRUE)
#compare the value with the version where na is omitted
c(median.steps.date,fill.median.steps.date)
```

## Are there differences in activity patterns between weekdays and weekends?  
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.  
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.  
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.  
```{r,echo=TRUE}
# create a new factor in the filled dataset with two levels "weekday" and "weekends"
# create the factor object by assuming that "Saturday" and "Sunday" as weekend, and the rests as weekdays
days<-weekdays(date.time)
days.index<-(unique(weekdays(date.time)))
days[days %in% days.index[1:5]]<-"Weekdays"
days[days %in% days.index[5:7]]<-"Weekend"
# create new factor.data by append days to fill.dataset
factor.data<-cbind(fill.data,days)
# split into weekdats and weekend data frame
weekdays.data<-split(factor.data,factor.data$days)$Weekdays
weekend.data<-split(factor.data,factor.data$days)$Weekend
# average number of steps taken each day across all weekdays of filled weekdays dataset
weekdays.interval.steps.date<-as.numeric(lapply(split(weekdays.data$steps,weekdays.data$interval),mean,na.rm=TRUE))
# average number of steps taken each day across all weekend of filled weekend dataset
weekend.interval.steps.date<-as.numeric(lapply(split(weekend.data$steps,weekend.data$interval),mean,na.rm=TRUE))
# construct a layout for the plot1
par(mfrow=c(2,1))
# construct histogram of average number of steps taken each day across all weekdays of filled weekdays dataset
plot(weekdays.interval.steps.date,type="l",main="Average Daily Number of Steps Taken in 5 Minutes Interval-Weekdays",xlab="Interval",ylab="Steps")
# construct histogram of average number of steps taken each day across all weekend of filled weekend dataset
plot(weekend.interval.steps.date,type="l",main="Average Daily Number of Steps Taken in 5 Minutes Interval--Weekend",xlab="Interval",ylab="Steps")
  
